---
title: "Probability"
author: "Jarod G.R. Meng"
date: "2/14/2021"
output:
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
    df_print: paged
  pdf_document: default
---

# Probability

## Probability distributions

### Normal distribution {#normal_dist}

#### Definitions

A **normal** (or **Gaussian**) distribution is a *continuous* probability distribution.

The general form of its **probability density function** is:

$$
f(x) = \frac{1}{\sigma\sqrt{2\pi}}e^{-\frac{1}{2}(\frac{x-\mu}{\sigma})^2}
$$

The parameter $\mu$ is the expected value or mean and $\sigma$ is the standard deviation. The variance of the distribution is $\sigma^2$.

The normal distribution is often referred to as $N(\mu,\sigma^2)$.

The simplest case of a normal distribution is known as the *standard normal distribution*. This is a special case when $\mu=0$ and $\sigma=1$. It can be described by the following probability density function.

$$
\varphi(x) = \frac{1}{\sqrt{2\pi}}e^{-\frac{1}{2}x^2}
$$

The shape of the probability density function is a **symmetric (around the mean) bell curve**. Its symmetry comes from the $x^2$ term. The negative sign in $-\frac{1}{2}x^2$ makes the curve greater in the middle and smaller on the sides.

```{r include=FALSE}
library(tidyverse)
theme_set(theme_minimal())
```


```{r}
ggplot() +
  xlim(-3, 3) +
  geom_function(fun = dnorm) +
  labs(
    x = expression(X %~% N(0, 1)), y = "probability density",
    title = "Normal distribution P.D.F"
  )
```

The **cumulative distribution function** of the standard normal distribution, usually denoted with the capital Greek letter $\Phi$, is the integral.

$$
\Phi(x) = \frac{1}{\sqrt{2\pi}}\int_{-\infty}^{x}e^{-t^2/2}dt
$$

```{r}
ggplot() +
  xlim(-3, 3) +
  geom_function(fun = pnorm) +
  labs(
    x = expression(X %~% N(0, 1)), y = "probability density",
    title = "Normal distribution C.D.F"
  )
```

The quantile function is the inverse of the cumulative distribution function. The quantile function of a standard normal distribution is called the **probit** function. It's denoted as $\Phi^{-1}(x)$.

```{r}
ggplot() +
  xlim(0, 1) +
  geom_function(fun = qnorm) +
  labs(
    x = "density", y = expression(X %~% N(0, 1)),
    title = "Normal distribution quantile function"
  )
```

The quantile $\Phi^{-1}(p)$ of the standard normal distribution is commonly denoted as $z_{p}$. These values are used in hypothesis testing, construction of confidence intervals, and Q-Q plots. In particular, $Z_{0.975}=1.96$ and $Z_{0.025}=-1.96$. They are often used to construct confidence intervals of 95% significant level.

#### Linear combination of two or more random variables

If $X_1$ and $X_2$ are two independent standard normal random variables with mean 0 and variance 1, then

1. Their sum and difference is distributed normally with mean 0 and variance 2

$$
X_1 \pm X_2 \sim N(0,2)
$$

1. If $X_1$, $X_2$, ..., $X_n$ are independent standard normal random variables, then the sum of their squares has the **chi-squared distribution** with $n$ degrees of freedom.

$$
X_1^2+...+X_n^2 \sim \chi_n^2
$$
