---
title: "rlang"
author: "Jarod G.R. Meng"
date: "5/1/2018"
output:
  md_document:
    variant: "markdown_github"
    toc: true
    toc_depth: 4
---

# R language

```{r setup, include=FALSE}
library(rlang)
knitr::opts_chunk$set(echo = TRUE)
```

## Concepts

* An R [expression][7] is an object that represents an *action* that can be
performed by R. It can be represented as a list of calls, symbols, constants
etc.

* [Quoting][1] or quotation is the action of capturing an expression instead of
evaluating it.


* **Quasiquotation** is the combination of quoting an expression while allowing
immediate evaluation (unquoting) of part of that expression.

  * [Unquoting][2] or unquotation allows selective evaluation parts of a quoted
  expression.

* Quosure

## Expressions

### Overview

An expression is also called an **abstract syntax tree** (**AST**) because it
represents the hierarchical tree structure of the code.

Note that the [Expressions][6] chapter in [Advanced R][9] refers to base R's
expression as [**expression objects**][12] and uses the term *expression* to
include following objects.

1. **Constants** are “self-quoting” in the sense that the expression used to
represent a constant is the constant itself.

1. [Symbols][11] represent variable names.

1. [Calls][10] define the tree in AST. A call behaves similarly to a
(potentially nested) list.

    * It has a `length()`.

    * You can extract elements with `[[`, `[`, and `$`.

    * Calls can contain other calls (hence nested list).

1. **Pairlists** have been replaced by lists almost everywhere. The only place
you are likely to see pairlists in R is when working with function arguments.

### Two ways of creating R expressions

1. You can **build** expressions using calls and symbols from parts and pieces,
e.g., using `call()` function in base R, or `call2()` in `rlang`.

1. The other way to create expressions is by quotation or quasiquotation, i.e.,
by intercepting an expression instead of evaluating it.

### Base R expression functions

#### Expression and its components

* `expression()` captures its arguments (unevaluated) in a vector of type
"expression" which can be coerced into a list.

  ```{r base_r_expression}
  x <- expression("R", a <- 1, 2 * 5, LETTERS)
  
  x
  as.list(x)
  typeof(x)
  class(x)
  length(x)
  sapply(x, typeof)
  sapply(x, class)
  ```

* `as.symbol()` and `as.name()` coerce the argument to a name.

  ```{r base_r_symbol}
  x <- as.symbol("R")
  
  typeof(x)
  class(x)
  ```

* `call()` creates an unevaluated function call which consists of the named
function applied to the given arguments.

  ```{r base_r_call}
  x <- call("*", 2, 5)
  
  x
  typeof(x)
  class(x)
  ```

  Note that although the returned call is unevaluated, the arguments supplied to
  `call()` are evaluated.
  
  ```{r base_r_call_args}
  a <- 2
  b <- 5
  x <- call("+", a, b)
  
  x
  ```

### Expression-related functions in `rlang`

* `expr()` and `exprs()` capture the expressions *you* supply (e.g., through
typing in an interactive session).

* `enexpr()` and `exexprs()` capture the expressions *user* supplies (e.g.,
through arguments in a function).

  * `ensym()` and `ensyms()` are variants of `enexpr()` and `enexprs()` that
check the captured expression is either a string (which they convert to symbol)
or a symbol. If anything else is supplied they throw an error.

#### `rlang` expression functions vs. base R expression functions

* `enexpr(arg)` corresponds to `substitute(arg)`.

* `expr()` is like `quote()`.

* `exprs()` is equivalent to `alist()`.

* `enexprs()` has no equivalent function in base R, but its behavior can be
reproduced with `eval(substitute(alist(...)))`.

The difference is that all quotation functions in `rlang` support unquotation,
so they are quasiquotation functions.

#### Conversion between string and expression

Conversion from strings to expressions is the base for functions like `source()`
.

* `parse()` returns the parsed but unevaluated expressions in a list.

  ```{r base_r_parse}
  x <- parse(
    text = c('"R"', "a <- 1", "2 * 5", "LETTERS")
  )
  
  x
  typeof(x)
  class(x)
  length(x)
  sapply(x, typeof)
  sapply(x, class)
  ```

* `deparse()` turns unevaluated expressions into character strings.

  ```{r base_r_deparse}
  x <- parse(text = "a <- 1")[[1]]
  y <- quote(b <- 2)
  
  deparse(x)
  deparse(y)
  ```

## Quotation

### Base R

There are two ways of quoting in base R.

1. [*Formula*][5] is one of the most important quoting operators in R. It's
usually used for the specification of statistical models.

1. The other family of quoting operator in base R is `quote()`, and
`substitute()` functions (see [documentation][4]).

    * `quote()` simply returns its argument unevaluated as a call.

      ```{r base_r_quote}
      x <- quote(seq(1, 10, 1))
      
      typeof(x)
      class(x)
      ```

    * `substitute()` takes an R expression and an `env` argument. It returns a
    call in which any variables bound in `env` are substituted.

      ```{r base_r_substitute}
      e <- new.env()
      e$a <- 1
      x <- substitute(a * 10, env = e)
      
      x
      typeof(x)
      class(x)
      ```

The difference is that **a formula captures its execution environment with its
expression** whereas **a quoting function only captures its expression**.

### `rlang` quotation

[1]: http://dplyr.tidyverse.org/articles/programming.html#quoting
[2]: http://dplyr.tidyverse.org/articles/programming.html#unquoting
[3]: https://stat.ethz.ch/R-manual/R-patched/library/base/html/tilde.html
[4]: http://stat.ethz.ch/R-manual/R-devel/library/base/html/substitute.html
[5]: https://stat.ethz.ch/R-manual/R-patched/library/stats/html/formula.html
[6]: https://adv-r.hadley.nz/expressions.html
[7]: https://stat.ethz.ch/R-manual/R-devel/library/base/html/expression.html
[8]: https://stat.ethz.ch/R-manual/R-devel/library/base/html/parse.html
[9]: https://adv-r.hadley.nz/
[10]: https://stat.ethz.ch/R-manual/R-devel/library/base/html/call.html
[11]: https://stat.ethz.ch/R-manual/R-devel/library/base/html/name.html
[12]: https://adv-r.hadley.nz/expressions.html#expression-objects
